Modelling BCG vaccination in the UK

### 2019-07-29 rmds/09-model-fitting.Rmd
scaffold strenghts and weaknesses

Libbi was used for all model fitting.[@Murray2015] LibBi is a software package for state-space modelling and Bayesian inference. It uses a domain specifc language for model specification, which is then optimised and compiled to provide highly efficient model code. It focusses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC^2 methods for parameter estimation. All fitting algorithms are highly efficient and scalable across mutliple CPUs or GPUs. `RBI` and `RBI.helpers` were used to interface with LibBi from R.[@Funk:2019ud; @Funk:2019uw] `RBI.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked, with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo^2 (SMC^2). This algorithm is used for all dynamic model fitting in this thesis.

@@

When using an MCMC alogorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC^2 a proposal distribution is required to inform the MCMC sampler that is run during the rejuvernation step. By default if no proposal distribution is provided Libbi uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

@@

I chose to use, manual, independent proposal adaption method for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and diffult to debug. Finally, for SMC^2 efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC^2 is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvernation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

@@

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC^2 step. Parameter sensitivity measures such as PRCC must be calculated seperately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

7. SMC^2 model fitting with 2000 initial parameter particles. Particle rejuvernation was set to trigger when the effective sample size decreased below 25%, with 5 MCMC steps used each time.

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC^2 algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvernation steps. This resulted in spuriously tight priors. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and any conclusions drawn from them must be appropriately caveated.

@@

[@Murray2015] suggests that increasing the number of particles used in SMC may improve the quality of the model fit. Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that each rejuvernation steps also need to be stored in RAM. Attempts to increase the amount of avaialble RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of parameters to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC^2 algorithm focussing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to even poor model fits.

Varying the proposal size, rejuvernation thresold, and number of rejuvernation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer runtime could result in an improved fit to data with no other model changes. Alternatively this results may driven by the model being too complex for an SMC^2 approach to viable in this framework.

@@

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, and presented the results of this fitted model.

* Summary of work from chapter
* Method strengths and weaknesses
* Software strengths and weaknesses
* Strengths and weaknesses
* Further work
* Conclusions

@@

* Defined the disease transmission model from the previous chapter as a state-space model. 
* Outlined the observed data and defined an appropriate measurement model for this data. 
* Developed a model fitting pipeline based on SMC^2 to fit the previously defined state-space model to TB notification data.
* Discussed ad-hoc approaches used to improve model fit. 
* Evaluated the various scenarios discussed in the previous chapter using DIC and discussed the implications of these findings. 
* Presented the model fit to data from the best fitting scenario as established using the DIC. 
* Discussed the posterior distributions of the fitted parameters and compared these distributions to the prior distributions.
* Estimated the model sensitivity to parameter changes and discussed the impact that this sensitivity may have.

>>>>

Libbi was used for all model fitting.[@Murray2015] LibBi is a software package for state-space modelling and Bayesian inference. It uses a domain specifc language for model specification, which is then optimised and compiled to provide highly efficient model code. It focusses on full information model fitting approaches including: particle Markov chain Monte Carlo (PMCMC), and SMC-SMC methods for parameter estimation. All fitting algorithms are highly efficient and scalable across mutliple CPUs or GPUs. `RBI` and `RBI.helpers` were used to interface with LibBi from R.[@Funk:2019ud; @Funk:2019uw] `RBI.helpers` was also used to optimise the model fitting pipeline as detailed in the calibration section. As model fitting using `LibBi` is compute intensive a workstation was built, and overclocked, with these compute requirements in mind^[See these blog posts for details: https://www.samabbott.co.uk/post/building-an-rstats-workstation/, https://www.samabbott.co.uk/post/benchmarking-workstation-xgboost/, https://www.samabbott.co.uk/post/benchmarking-workstation-benchmarkme/]. All model fitting code is available on GitHub as an R package^[See here for details: https://github.com/seabbs/ModelTBBCGEngland].

@@

   * *Weighting:* Parameter particles are weighted using the marginal likelihood. In principle this could be computed exactly, but is most commonly estimated using a nested particle filter for each state particle (i.e as outlined in the previous section). For a subset of models a Kalman filter maybe used instead.[@Murray2015] The marginal likelihood may also be estimated using other partial information techniques such as approximate bayesian computation. In the case where a particle filter is used the full algorithm is known as Sequential Monte Carlo - Sequential Monte Carlo (SMC-SMC). This algorithm is used for all dynamic model fitting in this thesis.

@@

When using an MCMC alogorithm a proposal distribution is required to provide new parameter samples to evaluate. For SMC-SMC a proposal distribution is required to inform the MCMC sampler that is run during the rejuvernation step. By default if no proposal distribution is provided Libbi uses the prior distribution.[@Murray2015] The prior distribution can be an inefficient proposal distribution as it is likely to have a low acceptance rate (from the MCMC sampler).[@Murray2015] Having a low acceptance rates means that many more MCMC steps are required to generate a successful parameter sample. This results in slow mixing and computationally expensive MCMC steps may make model fitting intractable.

@@

I chose to use, manual, independent proposal adaption method for several reasons. Firstly, when developing this pipeline the approaches implemented in `rbi.helpers` produced multiple transient errors in other `rbi` and `rbi.helpers` code. Secondly, the resulting dependent proposal distribution was highly complex, slow to compute, and diffult to debug. Finally, for SMC-SMC efficient exploration of the proposal distribution is less important than when using MCMC alone as SMC-SMC is initialised with multiple samples from the prior distribution. This means that multiple local maximas can be efficiently explored regardless of the proposal distribution used. The MCMC rejuvernation step then serves to provide additional samples from these local maximas. Proposal adaption was only carried out for the main model scenario with all other scenarios using this proposal distribution.

@@

I have implemented PRCC using the `epiR` R package^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/test_sensitivity.R],[@EpiR] using the samples from the posterior distribution of the model calculated during the SMC-SMC step. Parameter sensitivity measures such as PRCC must be calculated seperately for each output time point. I calculated the PRCC for each fitted parameter, at the final time point with fitted data (2004), for overall TB incidence rates. These results are then summarised by plotting the absolute PRCC values, indicating the direction of correlation using colour^[Code: https://github.com/seabbs/ModelTBBCGEngland/blob/master/R/plot_sensitivity.R].

@@

7. SMC-SMC model fitting with 2000 initial parameter particles. Particle rejuvernation was set to trigger when the effective sample size decreased below 25%, with 5 MCMC steps used each time.

@@

The pipeline outlined above resulted in a poor fit to the observed data. The SMC-SMC algorithm had a low effective sample size for each iteration, and a low acceptance rate for particle rejuvernation steps. This resulted in spuriously tight priors. Ad-hoc calibration (detailed in the following section) failed to improve the quality of this fit or find a subset of the model - or parameters - that fit the observed data to an acceptable degree whilst remaining computationally feasible. The results presented in the following section are therefore preliminary and any conclusions drawn from them must be appropriately caveated.

@@

[@Murray2015] suggests that increasing the number of particles used in SMC may improve the quality of the model fit. Unfortunately as `LibBi` stores SMC particle paths in Random access memory (RAM) the number of particles was restricted. An additional limitation is that each rejuvernation steps also need to be stored in RAM. Attempts to increase the amount of avaialble RAM (64 GB) using a 500GB SWAP (virtual memory) drive increased the upper limit on the number of particles but gains from this were restricted due to thrashing^[Thrashing: https://en.wikipedia.org/wiki/Thrashing_(computer_science)]. A major cause of the high RAM requirements of the model was that `LibBi` stores all parameters defined as initial conditions across every time point in the model, regardless of the settings used. Attempting to reduce the RAM footprint by fitting a greater number of parameters allowed for a greater number of parameters to be used but also increased the model degrees of freedom and hence upped the required number of particles required by a greater amount than provided by fitting the parameter. Reducing the number of particles used to estimate the marginal likelihood allowed a greater number of particles to be used in the outer SMC step but this resulted in highly inaccurate estimates of the likelihood. These highly inaccurate likelihood estimates resulted in the SMC-SMC algorithm focussing on parameter sets that had a low likelihood estimate yet fit the data poorly - ultimately leading to even poor model fits.

Varying the proposal size, rejuvernation thresold, and number of rejuvernation steps showed some promise at improving the quality of the model fit but ultimately computational constraints limited how much progress could be made using this approach. It is possible that a much longer runtime could result in an improved fit to data with no other model changes. Alternatively this results may driven by the model being too complex for an SMC-SMC approach to viable in this framework.

@@

### Summary

In this chapter I have formulated the disease transmission model developed in the previous chapter as a state-space model, developed a model fitting pipeline to fit this model to observed data, discussed the approaches taken to try and improve the quality of the model fit, and presented preliminary results from the model fitting pipeline.

### Pipeline strengths and weaknesses

* This chapter showcased the use of SMC a highly optimised tool for fitting a complex models using plug and play approaches.

* Unfortunately the quality of the model fit was very poor and the complexity of `LibBi` makes understanding the root cause very difficult. It is very possible that the model developed in the last chapter is too complex to be fitted using this approach - at least without the use of several orders of magnitude greater compute resources (or compute time).[@Murray2015] However, it is also possible that even with these resources this pipeline may not have produced a high quality model fit as the model developed in the previous chapter was clearly much more complex than those envisioned by the developers of the software.[@Murray2015] `LibBi`, `rbi` and `rbi.helpers` have great potential as a standardised toolbox for modellers. However, in their current state they are difficult to use beyond relatively simple use cases. This difficulty is compounded by sporadic, inconsistent, documentation of both the underlying software and it's `R` interface. On top of this, both `LibBi` itself and the R libraries that support it have multiple apparent bugs that can be extremly diffult to debug due to the layered nature of the software. `rbi` and `rbi.helpers` are under development and it is likely that many of these issues will be dealt with over time. However, `LibBi` itself has not had a major release since 2016, the community around it is largely inactive, and it's complex nature makes it diffcult for newcomers to help towards its development.

### Comparison to other methods

* SMC-ABC
* PMCMC
* Iterated filtering
* Maximum likelihood

### Comparison to other software

There are several other software tools available within the `R` ecosystem for fitting infectious disease models. These include `pomp`, `easyABC`, `EpiModel`, and `odin`.

In developing the work presented here, and in the previous chapter, a simpler model was developed using `pomp`. This work was not included in this thesis as it was not sufficiently developed. However, it highlight similar issues with the `pomp` package to those observed when using `LibBi`. Whilst `pomp`'s documentation, stability, and testing were much improved over `LibBi` it had similar limitations when it came to complex models. This was to such an extent that I developed numerous helper functions to deal with both the input and output of `pomp` models^[idmodelr: https://github.com/seabbs/idmodelr]. `pomp` also offered only a limited subset of fitting techniques with a heavy focus on iterated filtering.

Each of the packages discussed has several issues, with all suffering either from a lack of continued development (excepting `pomp`) or maturity. This may mean that for highly complex models more work is needed on the available open-source tools

### Futher work

Whilst the results presented here are not encouraging it may be the case that with a greatly extended runtime a much better fitting model may be found. The first step to testing this is to rerun the pipeline using several orders of magnitude more rejuvernation steps - as these have no RAM overhead. Even if the resulting model fits are still poor this may indicate areas for improvement - in either the model fitting pipeline or the model itself. An alternative option is to recode the model developed in the last chapter into a more generic form (i.e C) and to attempt to fit the model using other techniques that are less compute intensive and more robust - such as SMC-ABC. A final option would be to reimplement the model in the form required by another modelling package - such as `pomp`. However, this would mean running the risk of again getting stuck within a framework that is difficult to debug and may not be working as expected.

### Conclusions

The model fitting pipeline developed here was theoretically robust, and highly reproducible, but in practise did not produce a high quality model fit. It is difficult to determine whether this was caused by the complexity of the model combined with the high compute requirements of SMC-SMC or if the software implementation itself was at fault. However, `LibBi`'s high barrier to entry and difficulty of use made both implementing the model, and assessing whether the model fitting was working as expected more difficult. It is likely that a more generic model implementation coupled with a less compute intensive fitting approach would produce more useful results. This work does still have some merit as it pushed both `LibBi` and SMC-SMC to it's limits helping to define what the limitations of this approach, and specialised modelling packages more generally, may be. It is possible that with an extended runtime model fits may become more reliable and hence more usable.

@@

* Defined the disease transmission model from the previous chapter as a state-space model. 
* Outlined the observed data and defined an appropriate measurement model for this data. 
* Developed a model fitting pipeline based on SMC-SMC to fit the previously defined state-space model to TB notification data.
* Discussed ad-hoc approaches used to improve model fit. 
* Evaluated the various scenarios discussed in the previous chapter using DIC and discussed the implications of these findings. 
* Presented the model fit to data from the best fitting scenario as established using the DIC. 
* Discussed the posterior distributions of the fitted parameters and compared these distributions to the prior distributions.
* Estimated the model sensitivity to parameter changes and discussed the impact that this sensitivity may have.
